Safe Shared Memory for Concurrent Algol

### 2017-10-12 proposal.tex
incorporate slightly revised version of Farzaneh's text

\documentclass{article}
\usepackage[tt=false]{libertine}
\usepackage{hyperref}
\usepackage{fullpage}

@@

@@

Farzaneh will write some general material here.

>>>>

\documentclass{article}
\usepackage[tt=false]{libertine}
\usepackage{hyperref}
\usepackage{mathpartir}
\usepackage{amsthm}
\usepackage{fullpage}

@@

\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}{Corollary}[theorem]
\newtheorem{lemma}[theorem]{Lemma}

@@

A proramming language is specified by explaining both its
\emph{structure} and its \emph{computational meaning}; these are
captured in the static and the dynamic semantics respectively. Armed
with such a definition, it is then possible to prove critical
correctness and safety properties both for a language as a whole, and
also for individual programs in this language.

One of these basic safety criteria is called ``type safety'', a
property which is usually factored into \emph{type preservation}
(programs do not change their type as they execute) and
\emph{progress} (any well-typed program can take a step of
computation, or is a value). These laws may seem quite basic, but
depending on the definition and sophistication of the language, they
can imply extremely strong semantic properties such as liveness,
memory safety, causality~\cite{atkey-mcbride:2013}, and even cost or
complexity bounds.

To give a very simple example of static and dynamic semantics,
consider the following arithmetic language, adapted
from~\cite{harper:2016}:

\newcommand\IsOk[1]{{#1}\ \mathsf{ok}}
\newcommand\IsVal[1]{{#1}\ \mathsf{val}}
\newcommand\Step[2]{{#1}\mapsto{#2}}
\newcommand\Plus[2]{\mathtt{plus} (#1, #2)}
\newcommand\Num[1]{\mathtt{num}[#1]}

\paragraph{Statics} We define the static semantics by specifying which
expressions are ``OK'' (well-typed):
\begin{mathpar}
  \inferrule{
  }{
    \IsOk{\Num{n}}
  }
  \and
  \inferrule{
    \IsOk{e_1}
    \\
    \IsOk{e_2}
  }{
    \IsOk{\Plus{e_1}{e_2}}
  }
\end{mathpar}

In a more sophisticated language, the static semantics would be given
by a much more complicated form of judgment than $\IsOk{e}$, as we
will see later on.

\paragraph{Dynamics}
The dynamics are given by a transition system, writing $\IsVal{e}$ to
mean that $e$ is a \emph{value}, and $\Step{e}{e'}$ to mean that the
program $e$ takes a step to $e'$.
\begin{mathpar}
  \inferrule{
  }{
    \IsVal{\Num{n}}
  }
  \and
  \inferrule{
    \Step{e_1}{e_1'}
  }{
    \Step{\Plus{e_1}{e_2}}{\Plus{e_1'}{e_2}}
  }
  \and
  \inferrule{
    \IsVal{e_1}
    \\
    \Step{e_2}{e_2'}
  }{
    \Step{\Plus{e_1}{e_2}}{\Plus{e_1}{e_2'}}
  }
  \and
  \inferrule{
    n_1 + n_2 = n_3
  }{
    \Step{
      \Plus{\Num{n_1}}{\Num{n_2}}
    }{
      \Num{n_3}
    }
  }
\end{mathpar}

At this point, we can prove the type safety theorem in two lemmas:
\begin{lemma}[Progress]
  If $\IsOk{e}$, then either $\IsVal{e}$ or $\Step{e}{e'}$ for some
  $e'$.
\end{lemma}
\begin{lemma}[Preservation]
  If $\IsOk{e}$ and $\Step{e}{e'}$, then $\IsOk{e'}$.
\end{lemma}

Of course, this was a very trivial example, but this methodology
scales up to problems of seemingly arbitrary complexity. For instance,
a type system inspired by Kripke semantics and the modal logic S5 has
been deployed in the ML5 language for distributed computing, enabling
one to develop distributed applications within a single unified
program~\cite{murphy-crary-harper:2008}. Concurrency primitives have
been added to functional languages such as
Facile~\cite{giacolone:1989}, Concurrent ML~\cite{reppy:1991,reppy:2007},
JOCaml~\cite{conchon-fessant:1999}, Concurrent
Haskell~\cite{jones-gordon-finne:1996} and Concurrent
Algol~\cite{harper:2016}. In another work, Balzer and Pfenning
augmented the session-typed linear logic with shared memory, using
shared channels and shared memory with an acquire-release discipline.
